---
title: "Exploratory Data Analysis of English Premier League Opening Weekend"
output:
  html_document: default
  html_notebook: default
---


In honor of the start of a new season for the English Premier League (EPL), I am putting together this exploratory data analysis of historical EPL data to see how teams typically do to start a season. 

```{r libraries}
# loading in the required packages
suppressWarnings(suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lubridate)
  library(magrittr)
  library(tidyquant)
  library(purrr)
}))
```

First, we will load in the data. All of the data used in this analysis is from www.football-data.co.uk. 

```{r read-files}
files <- list.files(path = "epl_results", full.names = TRUE)

raw_data <- map(files, read.csv)
```


Below is the column information provided by the website:
    Div = League Division
    Date = Match Date (dd/mm/yy)
    HomeTeam = Home Team
    AwayTeam = Away Team
    FTHG and HG = Full Time Home Team Goals
    FTAG and AG = Full Time Away Team Goals
    FTR and Res = Full Time Result (H=Home Win, D=Draw, A=Away Win) 
    HTHG = Half Time Home Team Goals 
    HTAG = Half Time Away Team Goals 
    HTR = Half Time Result (H=Home Win, D=Draw, A=Away Win)

Match Statistics (where available)
    Attendance = Crowd Attendance
    Referee = Match Referee
    HS = Home Team Shots
    AS = Away Team Shots
    HST = Home Team Shots on Target
    AST = Away Team Shots on Target
    HHW = Home Team Hit Woodwork
    AHW = Away Team Hit Woodwork
    HC = Home Team Corners
    AC = Away Team Corners
    HF = Home Team Fouls Committed
    AF = Away Team Fouls Committed
    HO = Home Team Offsides
    AO = Away Team Offsides
    HY = Home Team Yellow Cards
    AY = Away Team Yellow Cards
    HR = Home Team Red Cards
    AR = Away Team Red Cards
    
There are more columns provided in the raw data set that have to do with betting odds, however, we will remove them as they are not going to be used in this analysis. Additionally, only two seasons data has attendance recorded, so this will be removed. Looking through the data, all of the data sets for the 2000/2001 through the 2016/2017 seasons have all of tje match statistics listed above. The data sets prior to the 2000/2001 data set only have the general data and none of the match statistics.  Becasue of these differences in available data we will only use the general statistics for the first part of this analysis. 

```{r combine-data}
data_general <- map(raw_data, function(x){
  output <- x %>%
    filter(Date != "") %>%  #the csv files pulled in some extra rows, this line removes them
    mutate(Date = dmy(Date)) %>% #converting the Date column from a factor to date object
    select(Div:HTR) %>%
    mutate(season = ifelse(month(Date) > 7,
                                   year(Date) + 1,
                                    year(Date)))  #creating a reference date 
  return(output)
})

data <- do.call(rbind, data_general) %>% as.tibble()
```


```{r show-combined, results = "hide"}
data 
```

```{r output-show-combined}
# # A tibble: 8,360 x 11
#       Div       Date    HomeTeam       AwayTeam  FTHG  FTAG    FTR  HTHG  HTAG    HTR season
#    <fctr>     <date>      <fctr>         <fctr> <int> <int> <fctr> <int> <int> <fctr>  <dbl>
#  1     E0 1995-08-19 Aston Villa     Man United     3     1      H     3     0      H   1996
#  2     E0 1995-08-19   Blackburn            QPR     1     0      H     1     0      H   1996
#  3     E0 1995-08-19     Chelsea        Everton     0     0      D     0     0      D   1996
#  4     E0 1995-08-19   Liverpool Sheffield Weds     1     0      H     0     0      D   1996
#  5     E0 1995-08-19    Man City      Tottenham     1     1      D     0     1      A   1996
#  6     E0 1995-08-19   Newcastle       Coventry     3     0      H     1     0      H   1996
#  7     E0 1995-08-19 Southampton  Nott'm Forest     3     4      A     1     3      A   1996
#  8     E0 1995-08-19    West Ham          Leeds     1     2      A     1     0      H   1996
#  9     E0 1995-08-19   Wimbledon         Bolton     3     2      H     2     2      D   1996
# 10     E0 1995-08-20     Arsenal  Middlesbrough     1     1      D     1     1      D   1996
# # ... with 8,350 more rows
```

Now that the data from the csv files is all in one data frame, we can do some manipulation to get it into a more tidy format. 

```{r tidy-data, results = "hide"}
data_tidy <- data %>%
  gather(key = "venue", value = team, HomeTeam:AwayTeam) %>% 
  arrange(Date) %>%
  mutate_if(is.factor, as.character) %>%
  mutate(venue = ifelse(venue == "HomeTeam",
                        "Home",
                        "Away"),
         FTR = case_when(venue == "Home" & FTR == "H" ~ "W",
                         venue == "Home" & FTR == "A" ~ "L",
                         venue == "Away" & FTR == "H" ~ "L",
                         venue == "Away" & FTR == "A" ~ "W",
                         TRUE ~ FTR),
         HTR = case_when(venue == "Home" & HTR == "H" ~ "W",
                         venue == "Home" & HTR == "A" ~ "L",
                         venue == "Away" & HTR == "H" ~ "L",
                         venue == "Away" & HTR == "A" ~ "W",
                         TRUE ~ HTR),
         FTGF = ifelse(venue == "Home", FTHG, FTAG),  #Full Time Goals For
         FTGA = ifelse(venue == "Home", FTAG, FTHG),  #Full Time Goals Against
         HTGF = ifelse(venue == "Home", HTHG, HTAG),  #Half Time Goals For
         HTGA = ifelse(venue == "Home", HTAG, HTHG),  #Half Time Goals Against
         points_earned = case_when(FTR == "W" ~ 3,           #adding points
                                   FTR == "D" ~ 1,
                                   FTR == "L" ~ 0)) %>% 
  select(Div, season, Date, team, venue, FTR, FTGF, FTGA, HTR, HTGF, HTGA, points_earned) %>%
  group_by(season, team) %>%
  mutate(points = cumsum(points_earned)) %>% #calculating the number of points each team has through out the season
  ungroup()

data_tidy
```

```{r show-tidy-data}
# # A tibble: 16,720 x 13
# # Groups:   season, team [440]
#      Div season       Date        team venue   FTR  FTGF  FTGA   HTR  HTGF  HTGA points_earned points
#    <chr>  <dbl>     <date>       <chr> <chr> <chr> <int> <int> <chr> <int> <int>         <dbl>  <dbl>
#  1    E0   1996 1995-08-19 Aston Villa  Home     W     3     1     W     3     0             3      3
#  2    E0   1996 1995-08-19   Blackburn  Home     W     1     0     W     1     0             3      3
#  3    E0   1996 1995-08-19     Chelsea  Home     D     0     0     D     0     0             1      1
#  4    E0   1996 1995-08-19   Liverpool  Home     W     1     0     D     0     0             3      3
#  5    E0   1996 1995-08-19    Man City  Home     D     1     1     L     0     1             1      1
#  6    E0   1996 1995-08-19   Newcastle  Home     W     3     0     W     1     0             3      3
#  7    E0   1996 1995-08-19 Southampton  Home     L     3     4     L     1     3             0      0
#  8    E0   1996 1995-08-19    West Ham  Home     L     1     2     W     1     0             0      0
#  9    E0   1996 1995-08-19   Wimbledon  Home     W     3     2     D     2     2             3      3
# 10    E0   1996 1995-08-19  Man United  Away     L     1     3     L     0     3             0      0
# # ... with 16,710 more rows
```

To ensure that our tidying did not create any missing values, we can use the `summarise_all()` function.

```{r check-NA, results = "hide"}
data_tidy %>% 
  summarise_all(function(x) sum(is.na(x)))
```

```{r show-check-NA}
# # A tibble: 1 x 11
#     Div season  Date  team venue   FTR  FTGF  FTGA   HTR  HTGF  HTGA
#   <int>  <int> <int> <int> <int> <int> <int> <int> <int> <int> <int>
# 1     0      0     0     0     0     0     0     0     0     0     0
```

Now that we know that the data is in a tidy format, we can begin exploring the data. As an Arsenal fan, I think we should start by looking at how Arsenal has done each year. 

```{r arsenal-points-graph-1, fig.width=12, fig.height=7}
data_tidy %>%
  filter(team == "Arsenal",
         season < 2007) %>%
  ggplot(aes(Date, points)) +
  facet_wrap(~season, scales = "free") +
  geom_line() +
  theme_tq() +
  scale_x_date(date_breaks = "1 month", date_labels = "%m/%d") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 12))

```

```{r arsenal-points-graph-2, fig.width=12, fig.height=7}
data_tidy %>%
  filter(team == "Arsenal",
         season >= 2007) %>%
  ggplot(aes(Date, points)) +
  facet_wrap(~season, scales = "free") +
  geom_line() +
  theme_tq() +
  scale_x_date(date_breaks = "1 month", date_labels = "%m/%d") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 12),
        axis.text.y = element_text(size = 12),
        strip.text = element_text(size = 12))

```

From these graphs, it looks like Arsenal follow a similar pattern every year, which is not surprising since the 2017 season was the first in Arsene Wenger's tenure that they have not finised in the top 4. Looking at the plot for the 2017 season, it is clear that the period of the season that killed their chances of finishing in the top 4 was the strech of games between Februrary and March where their point increase flat lined. Let's now take a look at team's average finishing point total.

```{r season-ending-stats}
# season_ending <- data_tidy %>%
  

```